Analyzing DECREASE trials for extent of data fabrication

### 2017-02-10 submission/manuscript.Rmd
Revise steps

Second, we meta-analyzed all studies combined, including a dummy predictor for the DECREASE and non-DECREASE studies to replicate results presented in Figure 4 of the 2014 meta-analysis [@bouri2014]. Surprisingly, our results showed more evidence for unequal subgroups than the original meta-analysis [@bouri2014]; whereas the original test for subgroup differences was $\chi^2(1)=3.91,p=.05$, our analyses indicated that $\chi^2(1)=`r round(mod_mods$QM, 2)`,p=`r round(mod_mods$QMp, 3)`$. Additionally, the original analyses showed substantial heterogeneity ($I^2=74.4$\%) where we found an $I^2=`r round(mod_mods$I2, 3)`$\%. Different heterogeneity estimators (e.g., `DerSimonian-Laird` instead of `REML`) did not resolve this difference. We tried to clarify these discrepancies by e-mailing the original authors, but failed to receive a response.

@@

We estimated the number of data points that would need to be fabricated to arrive at the estimates from the DECREASE studies, given that the non-DECREASE studies represent the true effect. This assumes, in contrast to Step 2, that the DECREASE studies in fact contain fabricated data. The estimates from Step 3 provide an indication of the extent of the potential data fabrication in the DECREASE studies under this assumption [@commissie2011;@commissie2012;@commissie2013;@bouri2014].

@@

In order to estimate the number of fabricated data points, we estimated the population effect distribution of mortality (in log odds) per condition in both the DECREASE- and non-DECREASE studies separately. For each of these, we ran a meta-analysis applying the same, verified methods used in Step 1. These separate analyses were necessary, considering that Step 1 only provided estimates of the overall risk ratio but not the odds conditional on exposure (beta-blocker or control) and study set (DECREASE- versus non-DECREASE). This resulted in four meta-analytic mortality estimates with an accompanying variance. Throughout the simulations, we use the point estimates to simulate genuine- and fabricated data, but supplement this by using the distribution estimates as sensitivity analyses.

We applied the inversion method to estimate the number of fabricated data points in the DECREASE-I and DECREASE-IV studies [@casella2002]. With the inversion method, we iteratively hypothesized that *X* out of *N* data points were fabricated (i.e., $0, 1, ..., N$) in the DECREASE-I and DECREASE-IV studies separately. For each study, we simulated 10,000 datasets per hypothesis of *X* fabricated data points and *N-X* genuine data points. For each simulated dataset (exact simulation procedure in next paragraph), we determined the likelihood of the results with 
\begin{equation}
\label{eq1}
L(\theta|\pi_{E},\pi_{C})=\pi_{E}^{n_{11}}(1-\pi_{E})^{n_{12}} \times \pi_{C}^{n_{21}}(1 - \pi_{C})^{n_{22}}
\end{equation}
where $\pi_{E}$ indicates the mortality rate of the experimental condition as drawn from the meta-analytic effect distribution ($\pi_{C}$ indicates the same for the control condition). The likelihood was computed under both the fabricated effect estimates (i.e., $L_{fabricated}$) and the genuine data (i.e., $L_{genuine}$). Table 2 indicates which cell sizes the various $n_{XX}$ refer to within the (simulated) data. We compared the likelihoods to determine whether the simulated data were more likely to arise from the genuine- ($L_{genuine}$) or from the fabricated data ($L_{fabricated}$). Note that using the likelihoods for this is a minor deviation from the preregistration, where we initially planned on using $p$-value comparisons ([osf.io/vnmzc](https://osf.io/vnmzc)).

@@

For each hypothesis of *X* out of *N* fabricated datapoints, we computed the probability that the fabricated data are more likely than the genuine data ($p_F=P(L_{fabricated}>L_{genuine})$).  Based on $p_F$, we computed the confidence interval for $X$ (i.e., $X_{LB};X_{UB}$). For a 95\% confidence interval, the lowerbound is equal to the $p_F$ closest to .025, whereas the upperbound is equal to the $p_F$ closest to .975.

We computed $p_F$ for all possibilities of *X* out of *N* fabricated datapoints in 10,000 randomly generated datsets, which were generated in three steps. For each dataset we:

1. Sampled (without replacement) *X* fictitious participants that would be the result of data fabrication (sampled across all conditions).

@@

Based on the meta-analytic effect from 2 and the cell sizes from 3, we computed the  likelihoods of $L_{fabricated}$ and $L_{genuine}$ using Equation \ref{eq1}. As mentioned before, $p_F$ was computed as the probability that the data were more likely under the estimates resulting from the (allegedly) fabricated data (i.e., the DECREASE trials) than under the estimates resulting from the genuine data (i.e., the non-DECREASE trials; $p_F=P(L_{fabricated}>L_{genuine})$).

@@

For DECREASE-I, the 95\% confidence interval for the estimated number of fabricated data points is [`r which.min(res_i_certain - .025) - 1` - `r which.max(res_i_certain - .975) - 1`] or [`r which.min(res_i_uncertain - .025) - 1` - `r which.max(res_i_uncertain - .975) - 1`] when based on a point estimate or a more uncertain distribution estimate, respectively. The left column of Figure 3 depicts the $p_F$ per *X* fabricated data points (top panel) and the bounds of the confidence interval when the degree of confidence is altered (lower panel). Staying clearly between the dotted lines in the top panel, depicting the 95\% CI (top: .975; bottom: .025), it becomes apparent that the degree of uncertainty is too high to make any estimates with a high degree of confidence. This due to the small sample size (i.e., $N=`r length(res_i_certain) - 1`$) combined with the availability of just the summary results. Only when the degree of confidence is lowered to around 75\% does the interval not span the entire sample size. As such, based on the summary results, little can be said about the extent of the data fabrication that occurred in the DECREASE-I trial, affirming the conclusions of the original committee report [@commissie2013].

For DECREASE-IV, the 95\% confidence interval for the estimated number of fabricated data points is [`r which.min(res_iv_certain - .025) - 1` - `r which.max(res_iv_certain - .975) - 1`] or [`r which.min(res_iv_uncertain - .025) - 1` - `r which.max(res_iv_uncertain - .975) - 1`] when based on a point estimate or a more uncertain distribution estimate, respectively. The minor difference between the estimates indicates that there is a high degree of confidence that data fabrication did occur. Nonetheless, the range of potentially fabricated data points is still estimated at approximately 1000; this indicates that the summary results are insufficient to provide more than an estimated lowerbound. However, this does indicate that it is possible not all data were fabricated (i.e., $N=`r length(res_iv_certain) - 1`$), increasing the importance of data provenance to discern between genuine and falsified data. This affirms the conclusion of the scientific integrity committee that the results of the DECREASE-IV trial are "scientifically incorrect" (p.11) [@commissie2012].

The results also highlight that, despite the lack of availability of raw data, summary results from larger samples result in more certainty about the estimated number of fabricated data points using the inversion method. This is due to decreased standard errors at the population level of the effects, providing less variation in the simulations and resulting in higher sensitivity. However, this also highlights that in order to prevent detection it would be in the fabricators' interest to fabricate small studies (assuming the fabricator wants to remain undetected).

>>>>

Second, we meta-analyzed all studies combined, including a dummy predictor for the DECREASE and non-DECREASE studies to reproduce results presented in Figure 4 of the 2014 meta-analysis [@bouri2014]. Surprisingly, our results showed more evidence for unequal subgroups than the original meta-analysis [@bouri2014]; whereas the original test for subgroup differences was $\chi^2(1)=3.91,p=.05$, our analyses indicated that $\chi^2(1)=`r round(mod_mods$QM, 2)`,p=`r round(mod_mods$QMp, 3)`$. Additionally, the original analyses showed substantial heterogeneity ($I^2=74.4$\%) where we found an $I^2=`r round(mod_mods$I2, 3)`$\%. Different heterogeneity estimators (e.g., `DerSimonian-Laird` instead of `REML`) did not resolve this difference. We tried to clarify these discrepancies by e-mailing the original authors, but did not receive a response.

@@

We estimated the number of data points that would need to be fabricated to arrive at the estimates from the DECREASE studies, given that the non-DECREASE studies represent the true population effect distribution. This assumes, in stark contrast to Step 2, that the DECREASE studies in fact contain fabricated data. The estimates from Step 3 provide an indication of the extent of potential data fabrication in the DECREASE studies [@commissie2011;@commissie2012;@commissie2013;@bouri2014].

@@

In order to estimate the number of fabricated data points, we estimated the population effect distribution of mortality (in log odds) per condition in both the DECREASE- and non-DECREASE studies. For each of these, we ran a meta-analysis applying the same, verified methods used in Step 1. These separate analyses were necessary, considering that Step 1 only provided estimates of the overall risk ratio but not the odds conditional on exposure (beta-blocker or control) and trial set (DECREASE- versus non-DECREASE). This resulted in four meta-analytic mortality estimates with an accompanying effect variance. Throughout the simulations, we use the point estimates to simulate genuine- and fabricated data but supplement this by using the more uncertain distribution estimates as sensitivity analyses.

We applied the inversion method to estimate the number of fabricated data points in the DECREASE-I and DECREASE-IV trials [@casella2002]. With the inversion method, we iteratively hypothesized that *X* out of *N* data points were fabricated (i.e., $X={0, 1, ..., N}$). For each combination of *X* and trial, we simulated 10,000 datasets. Each simulated dataset contained *X* fabricated data points and *N-X* genuine data points. For each simulated dataset (exact simulation procedure in next paragraph), we determined the likelihood of the results with 
\begin{equation}
\label{eq1}
L(\theta|\pi_{E},\pi_{C})=\pi_{E}^{n_{11}}(1-\pi_{E})^{n_{12}} \times \pi_{C}^{n_{21}}(1 - \pi_{C})^{n_{22}}
\end{equation}
where $\pi_{E}$ indicates the mortality rate of the experimental condition as drawn from the meta-analytic effect distribution ($\pi_{C}$ indicates the same for the control condition). The likelihood was computed under both the fabricated effect estimates (i.e., $L_{fabricated}$) and the genuine data (i.e., $L_{genuine}$). Table 2 indicates which cell sizes the various $n_{XX}$ refer to within the (simulated) data. Subsequently, we compared the likelihoods to determine whether the simulated data were more likely to arise from the genuine trials ($L_{genuine}$) or from the fabricated trials ($L_{fabricated}$). Note that using the likelihoods for this is a minor deviation from the preregistration, where we initially planned on using $p$-value comparisons ([osf.io/vnmzc](https://osf.io/vnmzc)).

@@

For each hypothesis of *X* out of *N* fabricated data points, we computed the probability that the fabricated data are more likely than the genuine data ($p_F=P(L_{fabricated}>L_{genuine})$).  Based on $p_F$, we computed the confidence interval for $X$ (i.e., $X_{LB};X_{UB}$). For a 95\% confidence interval, the lowerbound is equal to the $p_F$ closest to .025, whereas the upperbound is equal to the $p_F$ closest to .975.

We computed $p_F$ for all *X* out of *N* fabricated datapoints in 10,000 randomly generated datsets, which were generated in three steps. For each dataset we:

1. Sampled (without replacement and across conditions) *X* fictitious participants that would be the result of data fabrication.

@@

Based on the meta-analytic effect from 2 and the cell sizes from 3, we computed the  likelihoods of $L_{fabricated}$ and $L_{genuine}$ using Equation \ref{eq1}. As mentioned before, we computed $p_F$ as the probability that the data were more likely under the estimates resulting from the (allegedly) fabricated data (i.e., the DECREASE trials) than under the estimates resulting from the genuine data (i.e., the non-DECREASE trials; $p_F=P(L_{fabricated}>L_{genuine})$).

@@

For DECREASE-I, the 95\% confidence interval for the estimated number of fabricated data points is [`r which.min(res_i_certain - .025) - 1` - `r which.max(res_i_certain - .975) - 1`] or [`r which.min(res_i_uncertain - .025) - 1` - `r which.max(res_i_uncertain - .975) - 1`] when based on a point estimate or a more uncertain distribution estimate, respectively. The left column of Figure 3 depicts the $p_F$ per *X* fabricated data points (top panel) and the bounds of the confidence interval when the degree of confidence is altered (lower panel). Staying clearly between the dotted lines in the top panel, depicting the 95\% CI (top: .975; bottom: .025), it becomes apparent that the degree of uncertainty is too high to make any estimates about fabricated data points with a high degree of confidence. This is partly due to the small sample size (i.e., $N=`r length(res_i_certain) - 1`$) and the availability of just the summary results. Only when the degree of confidence is lowered to around 75\% does the interval not span the entire sample size. As such, based on the summary results, little can be said about the extent of the data fabrication that occurred in the DECREASE-I trial, affirming the conclusions of the original committee report [@commissie2013].

For DECREASE-IV, the 95\% confidence interval for the estimated number of fabricated data points is [`r which.min(res_iv_certain - .025) - 1` - `r which.max(res_iv_certain - .975) - 1`] or [`r which.min(res_iv_uncertain - .025) - 1` - `r which.max(res_iv_uncertain - .975) - 1`] when based on a point estimate or a more uncertain distribution estimate, respectively. The minor difference between the estimates indicates that there is a high degree of confidence that data fabrication did occur based on the difference of the results alone. Nonetheless, the range of potentially fabricated data points is still estimated at approximately 1000; this indicates that the summary results are insufficient to provide more than an estimated lowerbound. This does indicate that it is possible not all data were fabricated (i.e., $N=`r length(res_iv_certain) - 1`$), increasing the importance of well-documented data provenance to discern between genuine and falsified data. This affirms the conclusion of the scientific integrity committee that the results of the DECREASE-IV trial are "scientifically incorrect" (p.11) [@commissie2012].

The results also highlight that, despite the lack of availability of raw data, summary results from larger samples result in more certainty about the estimated number of fabricated data points using the inversion method. This is due to decreased standard errors at the population level of the effects, providing less variation in the simulations and resulting in higher sensitivity. However, this also highlights that in order to prevent detection it would be in the fabricators' interest to fabricate small studies (assuming the fabricator wants to remain undetected). This adds another factor to the unreliability of studies with small sample sizes.
